---
title: "Konza Vegetation Analysis"
author: "Samantha Cofer"
format: html
execute: 
  echo: false
  warning: false
  messege: false
---

## Introduction 

This analysis will look at vegetaion data from the konza National Ecological 
Observatory Network (NOEN) site located near Manhattan, KS. It will use
tidyverse packages to wrangle and visualize some of the data.

## Loading the data

The data are provided in csv files. We'll use the tidyverse package 'readr' 
to open into R. 

```{r}
#| label: read-csv

# load packages
library("readr")
library("dplyr")
library("ggplot2")
library("ggthemes")
library("knitr")

# read in the data
konza_data <- read_csv("data/NEON.D06.KONZ.DP1.10058.001.div_1m2Data.2025-08.basic.20251222T234053Z.csv")
```

## Data summarizing with `dplyr`

We'll next use the dplyr package to summarize some different apsects of the data.

```{r}
#| label: summarize-with-dplyr

# first summary to make will be a count of the number of observations in 
# each family
konza_data %>% 
  group_by(family) %>%
  tally() %>%
  arrange(desc(n)) %>%
  kable()
```

This analysis showed that Poaceae and Asteraceae are among the most common family
found in these plots. 

```{r}
#| label: mean-percent-cover-by-family

# The goal is to calculate the mean percent cover of each family 
konza_data %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover)) %>%
  kable()
```

This analysis showed that the largest percent cover is the `NA` taxa, which means
it likely wasn't able to be identified, but the highest mean percent cover is at 16%
across the plots was Fagaceae.

```{r}
#| label: filter-by-elevation

# The goal is to calculate the mean percent cover of each family 
konza_data %>%
  filter(elevation>390) %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover)) %>%
  kable()
```

Next we'll learn to use `mutate()`, which is a function to create a new column based on contents form an existing column.

```{r}
#| label: use-mutate

# select pulls out the columns you want
# mutate creates a mew column with whatever content you want
# which can include manipulation of an existing column
konza_data %>%
  select(scientificName, family, percentCover) %>%
  mutate(propCover = percentCover / 100) %>%
  top_n(20) %>%
  kable()
```


Next up, we'll use `ggplot2` to make some figures 

```{r}
#| label: histogram-percent-cover

konza_data %>%
  filter(!is.na(family)) %>%
  ggplot(aes(x = percentCover)) +
  geom_histogram() +
  labs(title = "Histogram of Percent Cover, All Species",
     x = "Percent Cover") +
  theme_bw()
```

## Next steps 

1. follow up on the really high percent cover values. Maybe say >25%?

Looks like it's mostly soil and dead plant matter.

The following plot shows how the percent cover of Poaceae changes, or doesn't 
change, with elevation differences. This is visualized to see how elevation
impacts the presence of Poaceae. 

```{r}
#| label: poaceae-percent-cover-by-elevation

konza_data %>%
  filter(!is.na(family)) %>%
  filter(family == "Poaceae") %>%
  ggplot() +
  geom_boxplot(aes(x = elevation,
                   y = percentCover,
                   group = elevation)) +
  labs(title = "Cover of Poaceae by Elevation",
       y = "Percent Cover (%)",
       x = "Elevation in Meters")
```

This shows that elevation does not significantly impact the presence of Poaceae.

Next, we will visualize the percent cover of Fabaceae by elevtion. This is to 
analyze how elevation impacts the prevelence of Fabacea through a smooth best-
fit line on a point graph.

```{r}
#| label: comparison-of-fabaceae-by-elevation


#| label: comparison-of-fabaceae-by-elevation
konza_data %>%
  filter(!is.na(family)) %>%
  filter(family == "Fabaceae") %>%
  ggplot(aes(x = elevation,
           y = percentCover,
           color = family)) + 
  geom_point() +
  geom_smooth() +
  labs(title = "Compariosn of Fabaceae by Elevation",
       y = "Percent Cover (%)",
       x = "Elevation in Meters")
```

This shows that Fabacea is most prevelent at elevations around 400m. 

Next, we will visualize family elevation vs. percent cover in
a point graph. This is to create an easily readable color coded graph to see 
which families are more prevelent at different elevations.

```{r}

#| label: family-elevation-percent-cover 

  ggplot(data = konza_data %>%
           filter(!is.na(family)),
       aes(x = elevation,
           y = percentCover)) +
  geom_point(aes(color = family)) +
  labs(title = "Percent Cover by Family",
       y = "Percent Cover (%)",
       x = "Family")
  
```

After filtering out the NA families, which was a majority of high percent 
covers, we can clearly see that most families are under 25% cover in most samples.
