---
title: "Konza Vegetation Analysis"
author: "Samantha Cofer"
format: html
---

## Introduction 

This analysis will look at vegetaion data from the konza National Ecological 
Observatory Network (NOEN) site located near Manhattan, KS. It will use
tidyverse packages to wrangle and visualize some of the data.

## Loading the data

The data are provided in csv files. We'll use the tidyverse package 'readr' 
to open into R. 

```{r}
#| label: read-csv

# load packages
library("readr")
library("dplyr")
library("ggplot2")

# read in the data
konza_data <- read_csv("data/NEON.D06.KONZ.DP1.10058.001.div_1m2Data.2025-08.basic.20251222T234053Z.csv")
```

## Data summarizing with `dplyr`

We'll next use the dplyr package to summarize some different apsects of the data.

```{r}
#| label: summarize-with-dplyr

# first summary to make will be a count of the number of observations in 
# each family
konza_data %>% 
  group_by(family) %>%
  tally() %>%
  arrange(desc(n))
```

This analysis showed that Poaceae and Asteraceae are among the most common family
found in these plots. 

```{r}
#| label: mean-percent-cover-by-family

# The goal is to calculate the mean percent cover of each family 
konza_data %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover))
```

This analysis showed that the largest percent cover is the `NA` taxa, which means
it likely wasn't able to be identified, but the highest mean percent cover is at 16%
across the plots was Fagaceae


